---
title: "SignalCell"
geometry: margin=0.5cm
author: "Kyle Gellatly"
output:
  html_document:
    toc: true
    toc_float : yes
    toc_depth : 3
    theme : cerulean
---

```{r setup, include=FALSE, warning=FALSE, error=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
# https://github.com/kgellatl/SignalCell
library(SignalCell)
```

## Load Data

```{r}
# load("~/Dropbox (UMass Medical School)/UMASS/GitHub/R/SignalCell/data/tst_dat.Rdata")
# sparse_dat <- as(tst_dat, "dgCMatrix")

load("~/Dropbox (UMass Medical School)/Lab_inDrop/mDC_vignette_data/mDC_dat.Rdata")
sparse_dat <- as(mDC_dat, "dgRMatrix")
object.size(mDC_dat)
object.size(sparse_dat)
rm(mDC_dat)

```

# SCE Class

## Construct SingleCellExperiment

```{r}
### http://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html
sce <- construct_sce(sparse_dat)
rm(sparse_dat)

class(sce)
class(assay(sce, "counts"))
sce

```

### Enhanced exprs Like functionality

```{r}

counts <- assay(sce, "counts")
libsizes <- colSums(counts)
size.factors <- libsizes/mean(libsizes)
logcounts(sce) <- log2(t(t(counts)/size.factors) + 1)
assayNames(sce)

get_def_assay(sce)
sce <- set_def_assay(sce, "logcounts")
get_def_assay(sce)

```

### pData Like functionality

```{r}

colData(sce)$Timpoint <- "0hr"
sce$Timpoint[grep("1", colnames(sce))] <- "1hr"
sce$Timpoint[grep("4", colnames(sce))] <- "4hr"

sce <- calc_libsize(sce, assay = "counts", label = "UMI_sum")
colData(sce)
sce

```

### fData Like functionality

```{r}

counts <- apply(assay(sce, "counts"), 1, function(c)sum(c!=0))
rowData(sce)$gene_counts <- counts
rowData(sce) # use for vector annotations
sce

```

# Basic Analysis

## Finding variable genes

```{r}

sce <- calc_var_genes(sce, method = "Malhanobis")
sce <- calc_var_genes(sce, method = "CV")
sce <- calc_var_genes(sce, method = "Gini")

rowData(sce)
sce
```

## Bare Minimum Example

```{r}
gene_subset <- get_var_genes(sce, method = "Malhanobis", cutoff = 0.85)
gene_subset[100:110]
sce <- dim_reduce(input = sce, genelist = gene_subset, method = "iPCA")
sce <- embed_cells(input = sce, lem = "iPCA", method = "tSNE")
colData(sce)
```


### Dimension reduction Detailed

```{r}
###### Example 1 : Linear Embedding Matrix (LEM) Class, Single Embedding #####

# First we run dimension reduction
# When NULL, the name of the lem defaults to the method, in this case iPCA
sce <- dim_reduce(input = sce, genelist = gene_subset, method = "iPCA", reducedDim_key = NULL)

# We can see what LEM results exist with reducedDimNames
reducedDimNames(sce)

# We can access the LEM with reducedDim and providing a LEM name
LEM_iPCA <- reducedDim(sce, "iPCA")

# Structure of the LEM
# SVD (like) Matrices
head(sampleFactors(LEM_iPCA)) # cells = rows
head(featureLoadings(LEM_iPCA))  # gene = rows
head(factorData(LEM_iPCA)) # diagonal

# There is also a metadata field. This stores the parameters of the LEM.
# As of now, the embedding field is empty
metadata_iPCA <- metadata(LEM_iPCA)
str(metadata_iPCA)

# Next we embed the cells. Right now tSNE is supported, the default method. We can add UMAP support
# This embedding is stored internally to the input LEM, so a lem must be provided as an argument
# As above, when no embedding_key is provided, the name of the embedding defaults to the method, in this case tSNE
# When TRUE, write_colData will write the x & y coordinates to colData()
# When NULL the colData_labs default to x and y pasted with the method

sce <- embed_cells(input = sce, lem = "iPCA", embedding_key = NULL, write_colData = T, method = "tSNE", seed = 10)
colData(sce)

# Now within the metadata of the LEM, we have an embedding stored, along with its parameters
# To simplify usage, there is a function, embeddings(), to list available LEMs, and their embedding_key
# There is another function, embedding(), to retrieve a specific embedding with its embedding_key
# This requires a LEM argument, and the embedding_key

embeddings(sce)
head(embedding(sce, lem = "iPCA", embedding = "tSNE"))

# Parameters for the embedding are nested within the LEM metadata, and then the embedding
LEM_iPCA <- reducedDim(sce, "iPCA")
metadata_iPCA <- metadata(LEM_iPCA)
metadata_iPCA$embeddings$tSNE$parameters

######  Example 2 : Second LEM, Customization, multiple embeddings ######

# You can store as many LEM objects you want.
# In this case we use reducedDim_key to name it specifically.
# We run PCA for 20 components
sce <- dim_reduce(input = sce, reducedDim_key = "PCA_20comp", genelist = gene_subset, method = "PCA", nComp = 20)

# Now we have 2 LEMs stored
reducedDimNames(sce)

# Now when we call embed_cells, we call the new reducedDim() object, "PCA_20comp"
# However we only select 15 components
# When embedding we can also specify the column names written to colData()
# This way as many x + y embeddings can be stored as you desire within colData (as long as they are uniquely named)
sce <- embed_cells(input = sce, lem = "PCA_20comp", embedding_key = "PCA_20_15", colData_labs = c("PCA_20_15_x", "PCA_20_15_y"), comp_select = 1:15, iterations = 500, tSNE_perp = 50)
colData(sce)

# On a single LEM object we can try as many embeddings as desired
# To prevent them overwriting, we can provide a new embedding_key
# With our second embedding, we tweak parameters
sce <- embed_cells(input = sce, lem = "PCA_20comp", embedding_key = "PCA_embed2", colData_labs = c("PCA_embed2_x", "PCA_embed2_y"), comp_select = 1:10, tSNE_perp = 50)
colData(sce)

# Now our object has 2 sets of LEMS, and within the second LEM are 2 different embeddings
embeddings(sce)
```

## Clustering

```{r}
reducedDims(sce)
sce <- cluster_cells(sce, dims = "Comp", lem = "iPCA",  method = "spectral", k = 6)

# 2D clustering can  be done on colData()
sce <- cluster_cells(sce, dims = "2d", method = "density", xy = c("tSNE_x", "tSNE_y"))
colData(sce)

# 2D clustering can also be done on an embedding not stored in colData, but requires more arguments because the embedding is stored with the LEM
# We also provide a name arguement for this new colData() column to prevent overwriting the default "Cluster"
# Notice the end result clustering on the colData() or lem are the same, because the underlying data is the same
embeddings(sce)
sce <- cluster_cells(sce, dims = "2d", method = "density", lem = "iPCA",  embedding = "tSNE", name = "PCA_20_embed2_cluster")
colData(sce)[,c("Cluster", "PCA_20_embed2_cluster")]

# If write_colData is FALSE when you run embed_cells(), than the colData will not get flooded with all embeddings you run
# You can always write a single embedding result to colData, the one you prefer
int_embedding <- embedding(input = sce, lem = "iPCA", embedding = "tSNE")
colnames(int_embedding) <- c("int_embedding1", "int_embedding2")
colData(sce) <- cbind(colData(sce), int_embedding)
colData(sce)
```

# Basic Plots

## Setting default coordinates

```{r}
###### NEEDS TO BE CHANGED
#### EITHER EMBEDDING OR LEM!
# sce <- set_xy(sce, "iPCA", x = 1, y = 2)
# head(colData(sce))[,c("x", "y")]
# 
# sce <- set_xy(sce, "iPCA")
# head(colData(sce))[,c("x", "y")]

```

## tSNE metadata Plot

```{r}

###### NEEDS TO BE CHANGED
#### EITHER EMBEDDING OR LEM!

# plot_metadata(sce, color_by = "Cluster")
# 
# plot_metadata(sce, color_by = "Cluster", facet_by = "Timepoint",  coords = c("iPCA", 1, 2))

```

## tSNE gene Plot

```{r}

###### NEEDS TO BE CHANGED
#### EITHER EMBEDDING OR LEM!

# plot_tsne_gene(sce, gene = "Ccr7")
# 
# plot_tsne_gene(sce, gene = c("Ccr7", "Lcn2")) 
# 
# plot_tsne_gene(sce, gene = c("Tnf"), facet_by = "Timepoint")

```

## Violin Plot

```{r}

# plot_violin(sce, gene = "Ccr7", color_by = "Cluster")
# 
# plot_violin(sce, gene = "Ccr7", color_by = "Cluster", assay = "counts")
# 
# plot_violin(sce, gene = c("Ccr7", "Lcn2"), color_by = "Cluster") # NEW
# 
# plot_violin(sce, gene = c("Tnf"), facet_by = "Timepoint", color_by = "Cluster")

```


# Bulk Analysis

```{r}

# sce@elementMetadata # use for bulk values??

```

# To Do

# Change the dim reduce

### Quantitative clustering

### Test file Sizes on LARGE Data! 

### dgC vs dgR matrix performance test

